---
title: "Model2_SodiumLinearRegression"
author: "Team 1"
date: "2025-12-12"
output:
  rmdformats::readthedown:
    number_sections: true
    toc_depth: 3
    toc_float: yes
    code_folding: hide
---

```{r init, include=F}
library(arrow)
library(ggplot2)
library(ezids)
library(labelled)
library(gtsummary)
library(dplyr)
library(tidyr)
library(broom)
library(knitr)
library(lattice)
library(corrplot)
library(car)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  results = "markup")
  options(scientiic=T, digits = 3)
  
# Set the root directory for all code chunks (important for reading files)
knitr::opts_knit$set(root.dir = "/Users/akashreddyashanna/Desktop/Fall 2025/Intro to DS/DataScienceProject")
```

# Reading the parquet file
```{r}
food <- read_parquet("proc_data/food_final_morevars.parquet")
str(food)
summary(food)
```

# Preprocessing
```{r}
# Subset with only focus variables
food_model <- subset(food, select = c(gender,age,race,timeus,hhsize,income_pov_cat,sodium))
str(food_model)
xkablesummary(food_model)

# Checking for missing values
sum(is.na(food_model))
food_model <-  food_model[!is.na(food_model$income_pov_cat),]
xkablesummary(food_model)
```

# Exploratory Data Analysis
## Histogram of Sodium distribution
```{r}
hsodium <- ggplot(food_model, aes(sodium)) +
  geom_histogram(
    breaks=seq(0, 21000, by = 1000),
    fill = "chocolate",
    color = "chocolate4",
    alpha = 0.9
  ) +
  labs(
    title = "Distribution of Daily Sodium Intake",
    x = "Total Sodium (mg)",
    y = "Frequency"
  ) +
  theme(
    # Background colors
    plot.background = element_rect(fill = "antiquewhite", color = NA),   # outside area
    panel.background = element_rect(fill = "antiquewhite", color = NA),  # plot area
    panel.grid.major = element_line(color = "white"),             # optional: lighten grid
    panel.grid.minor = element_blank(),
    
    # Font sizes
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),  # title
    axis.title = element_text(size = 20, face = "bold"),               # axis titles
    axis.text = element_text(size = 20)                                # axis labels
  )

# Display the plot
hsodium
```

**Sodium intake is heavily right-skewed which shows that most people consume moderate amounts, but there are few extreme outliers with very high intakes.**

## Boxplot of Sodium Intake by genders
```{r}
box_sodium_gender <- ggplot(food_model, aes(x = gender, y = sodium, fill = gender)) +
  geom_boxplot(
    alpha = 0.8, 
    color = "chocolate4"
  ) +
  scale_fill_brewer(
    palette = "Set3"
  ) +
  labs(
    title = "Boxplot of Sodium Intake by Genders",
    x = "Genders",
    y = "Sodium Intake (mg)"
  ) +
  theme(
    # Background colors
    plot.background = element_rect(fill = "antiquewhite", color = NA),
    panel.background = element_rect(fill = "antiquewhite", color = NA),
    panel.grid.major = element_line(color = "white"),
    panel.grid.minor = element_blank(), 
    
    #Font sizes
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 20, face = "bold"),
    axis.text  = element_text(size = 20),
    legend.position = "none"
  )
# Display the plot
box_sodium_gender
```

**This boxplot compares sodium intake by gender. We can see that Males have a higher median sodium intake and a wider range of consumption compared to Females.**

## Scatterplot of Sodium Intake by different ages
```{r}
scatter_sodium_age <- ggplot(food_model, aes(x = age, y = sodium)) +
  geom_point(
    alpha = 0.5, 
    color = "chocolate4"
    ) +
  scale_fill_brewer(
    palette = "Set3"
    ) +
  labs(
    title = "Scatterplot of Sodium Intake by different Ages",
    x = "Age",
    y = "Sodium Intake (mg)"
    ) +
  theme(
    # Background colors
    plot.background = element_rect(fill = "antiquewhite", color = NA),
    panel.background = element_rect(fill = "antiquewhite", color = NA),
    panel.grid.major = element_line(color = "white"),
    panel.grid.minor = element_blank(), 
    
    #Font sizes
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 20, face = "bold"),
    axis.text  = element_text(size = 20),
    legend.position = "none"
    )
# Display the plot
scatter_sodium_age
```

**This scatterplot of sodium intake versus Age shows that data is quite dense, revealing a lot of variability at every age, but it shows that consumption of sodium decreases as people get older.**

## Boxplot of Sodium Intake by different Ethnic Groups
```{r}
box_sodium_ethnic <- ggplot(food_model, aes(x = race, y = sodium, fill = race)) +
  geom_boxplot(
    alpha = 0.8, 
    color = "chocolate4"
  ) +
  scale_fill_brewer(
    palette = "Set3"
  ) +
  labs(
    title = "Boxplot of Sodium Intake by different Ethnic Groups",
    x = "Ethnic Groups",
    y = "Sodium Intake (mg)"
  ) +
  theme(
    # Background colors
    plot.background = element_rect(fill = "antiquewhite", color = NA),
    panel.background = element_rect(fill = "antiquewhite", color = NA),
    panel.grid.major = element_line(color = "white"),
    panel.grid.minor = element_blank(), 
    
    #Font sizes
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 20, face = "bold"),
    axis.text  = element_text(size = 8),
    legend.position = "none"
  )
# Display the plot
box_sodium_ethnic
```

**From this plot, we can clearly see that the Non-Hispanic Asian group has a higher median sodium intake compared to others, while Hispanic and White groups appear to have lower overall sodium levels.The variation also differs slightly that means some groups have wider spreads in sodium consumption, likely due to diverse diets within those groups.**

## Boxplot of Sodium Intake by different Income Groups
```{r}
box_sodium_income <- ggplot(food_model, aes(x = income_pov_cat, y = sodium, fill = income_pov_cat)) +
  geom_boxplot(
    alpha = 0.8, 
    color = "chocolate4"
  ) +
  scale_fill_brewer(
    palette = "Set3"
  ) +
  labs(
    title = "Boxplot of Sodium Intake by different Income Groups",
    x = "Income Groups",
    y = "Sodium Intake (mg)"
  ) +
  theme(
    # Background colors
    plot.background = element_rect(fill = "antiquewhite", color = NA),
    panel.background = element_rect(fill = "antiquewhite", color = NA),
    panel.grid.major = element_line(color = "white"),
    panel.grid.minor = element_blank(), 
    
    #Font sizes
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 20, face = "bold"),
    axis.text  = element_text(size = 15),
    legend.position = "none"
  )
# Display the plot
box_sodium_income
```

**This shows the relationship between income and sodium.In this income is categorized by the 'income-to-poverty ratio'.From the lowest income group (<100) on the left to the highest income group (500+) on the right, we observe a slight upward trend in median sodium intake. This suggests that higher socioeconomic status is actually associated with higher sodium consumption.**

## Allpairs plot 
```{r}
pairs_plot <- pairs(food_model,
      labels = colnames(food_model),
      lower.panel = points,
      upper.panel = panel.smooth,
      diag.panel = NULL,
      gap = 0.5,
      cex.labels = 1.2,
      font.labels = 2,
      pch = 19,
      col = "lightblue",
      main = "Pairs Plot of food_model")
```

**This pairwise plot checks for correlations between all the variables.The relationship with age shows a downward trend, confirming the earlier observation. The relationship with household size appears relatively flat, suggesting weak association. We don't see any perfect straight lines, which tells us that no single variable perfectly predicts sodium intake on its own. It also helps us confirm that our predictors aren't too heavily correlated with each other, which avoids multicollinearity issues.**

## Correlation Matrix and plot
```{r}
#Subset of only numerical variables
food_model_num <- food_model[,c("age","hhsize","sodium")]
str(food_model_num)

cor_matrix <- cor(food_model_num)
print(cor_matrix)
corr_plot <- corrplot(cor_matrix,
                        method = "color",
                        col = COL2('PiYG'),
                        type = "full",
                        addCoef.col = "black",
                        tl.col = "black",
                        tl.srt = 45,
                        diag = TRUE)

```

**Correlation matrix and plot with the subset to only numerical variables age, hhsize and sodium.We can see that age and sodium are negtively correlated while hhsize and sodium are positively correlated.**


# Linear Models

## Model-1 with only numerical variables (age and hhsize)
```{r}
model1 = lm(sodium ~ age+hhsize, data = food_model)
summary(model1)
xkabledply(model1, title = "Model for sodium with age and hhsize as regressors" )
xkablevif(model1)
```

**We chose age and hhsize as the initial predictors because they are key numerical demographic variables**

**Coefficient Interpretation**

- **Intercept = 3896.513**, which represents the expected daily sodium intake (in milligrams) when both age and household size are zero. While this scenario is not practically meaningful, the intercept serves as a baseline level of sodium intake against which the effects of age and household size are measured.

- **Age Coefficient = −13.605**, For every one-year increase in age, daily sodium intake decreases by approximately 13.6 mg on average, holding household size constant.This shows a negative linear relationship between age and sodium intake, suggesting that older individuals tend to consume less sodium than younger individuals.

- **hhsize Coefficient = −27.961**, For each additional household member, sodium intake decreases by approximately 28 mg on average, holding age constant.

**P-Values:**

- **Intercept p-value < 2e-16**. This is highly statistically significant, indicating strong evidence that the baseline sodium intake differs from zero.

- **Age p-value < 2e-16**.This extremely small p-value indicates that age is a highly significant predictor of sodium intake. We reject the null hypothesis that the age coefficient equals zero.

- **hhsize p-value = 0.138**. This p-value exceeds alpha levels (0.05), so we fail to reject the null hypothesis for household size. Thus, household size does not significantly predict sodium intake in this model.

**Mutiple R-squared = 0.018113 (1.81%)**. This model explains approximately 1.8% of the total variation in daily sodium intake. This indicates that age and household size alone capture only a small portion of the factors influencing sodium consumption.

**Adjusted R-squared = 0.01767**. The adjusted value confirms the limited explanatory power of the model after accounting for the number of predictors.

**F-statistic p-value < 2.2e-16**.The overall model is statistically significant, meaning that at least one predictor (in this case, age) is associated with sodium intake.

**This baseline model shows that age is a strong and statistically significant predictor of sodium intake, with older individuals consuming less sodium on average. However, household size does not have a significant effect, and the low R-squared value indicates that most of the variation in sodium intake is explained by other demographic and socioeconomic factors not included in this model.**


## Model-2 with all focus variables (including categorical) 
```{r}
model2 = lm(sodium ~ age+gender+race+income_pov_cat+timeus, data = food_model)
summary(model2)
xkabledply(model2, title = "Model for sodium with all focus variables")
xkablevif(model2)
```

**We chose Model 2 to expand upon Model 1 by incorporating key categorical demographic and socioeconomic variables such as age, gender, race, income category, and length of time in the U.S,to better explain variation in sodium intake.**

**Coefficient Interpretation**

- **Age Coefficient = −14.71**, For every one-year increase in age, daily sodium intake decreases by 14.71 mg on average, holding other variables constant.This shows a negative linear relationship between age and sodium intake, suggesting that older individuals tend to consume less sodium than younger individuals.

- **GenderFemale Coefficient = −862.96**,This indicates that females are predicted to consume approximately 863mg less sodium than males, holding other variables constant.This suggest substantial gender based differences in dietary sodium intake.

- **Race Non-Hispanic Asian Coefficient = 392.32**, This shows that these individuals consume significantly more sodium than other ethnic groups, holding other variables constant.

- **Income Categories Coefficient**,Individuals in the 300–400% poverty threshold and 500%+ categories have significantly higher sodium intake compared to lower income group, suggesting that higher income does not necessarily correspond to healthier sodium consumption patterns.

- **timeUS Coefficients**, people who have been in US for 20 years and more have higher sodium intake while people who have been in US for 5 to 15 years have a negative effect as the sodium intake decreases with one unit increase in timeUS. 

**P-Values:**

- **Age  and genderFemale p-value < 2e-16**.This extremely small p-value indicates that age and genderFemale are highly significant predictors of sodium intake. We reject the null hypothesis that the age and genderFemale coefficient equals zero.

- **Race Categories p-value** Categories like Non-Hispanic White (p = 0.0278) and Non-Hispanic Asia (0.0032) were statistically significant (p < 0.05), While Non-Hispanic Black (0.1803) and Other Race - Including Multi-Racial (p=0.1171) do not show statistically significant differences because p > 0.05.

- **Income_pov_cat p-values** Categories like income_pov_cat300-400 (p=3.1e-05) and income_pov_cat500+ (p=5.9e-05) were statistically significant (p < 0.05), while income_pov_cat100-200 (p = 0.2253), income_pov_cat200-300 (p = 0.0588) and income_pov_cat400-500 (p = 0.0704) were statistically not significant because p > 0.05.

- **timeus p-value = 0.9707**, this is not statistically significant, indicating that duration do not reliably predicts sodium intake in this model.

**Mutiple R-squared = 0.101 (1.01%)**. This model explains approximately 10.1% of the total variation in daily sodium intake. This model explains over five times more variance than Model 1, representing a substantial improvement in explanatory power.

**Adjusted R-squared = 0.0986**. The adjusted value accounts for the 13 predictors and confirms that the additional variables meaningfully contribute to the model.

**F-statistic p-value < 2.2e-16**.The overall model is statistically significant, meaning that the predictors, as a group, are associated with sodium intake.

**Model 2 demonstrates that gender, age, income_pov_cat, and certain ethnic groups are strong predictors of sodium intake. Including demographic and socioeconomic variables significantly improves model performance, though a large portion of variability remains unexplained.**

## Model-3 with age and income as interaction terms
```{r}
model3 = lm(sodium ~ age+gender+race*income_pov_cat, data = food_model)
summary(model3)
xkabledply(model3, title = "Model for sodium with all focus variables including interaction between race and income")
xkablevif(model3)
```

**The model3 introduces interaction effects between race and income, allowing the relationship between income and sodium intake to differ across racial/ethnic groups.**

**Coefficient Interpretation**

- **Main Effect Coefficients** The interpretation of main effects (Age, Gender) becomes more complex, as their effect is now dependent on the interaction terms.

- **Interaction Term Coefficients (race * income_pov_cat)** This generated 20 new coefficients. These coefficients represent the additional increase or decrease in sodium intake for a specific race within a specific income category, beyond what the two main effects would predict.

**P-Values:**

- **Age  and genderFemale p-value < 2e-16**.This extremely small p-value indicates that age and genderFemale are highly significant predictors of sodium intake. We reject the null hypothesis that the age and genderFemale coefficient equals zero.

- **Race and Income Categories p-value** Several ethnic categories like Non-Hispanic black, Non-Hispanic Asian and income categories like 300–400%, 400–500% and 500+ were statistically significant (p < 0.05), While others do not show statistically significant differences because p > 0.05.

- **Race * Income_pov_cat p-values** The majority of the individual interaction terms were not statistically significant (p > 0.05), meaning most specific race-by-income combinations do not have a unique effect.

**Mutiple R-squared = 0.108 (1.08%)**. This model explains approximately 10.8% of the total variation in daily sodium intake. This model explains the highest percentage of variation, a 0.6% increase over Model 2.

**Adjusted R-squared = 0.101**. The adjusted value accounts for these predictors and confirms that the additional variables meaningfully contribute to the model.

**F-statistic p-value < 2.2e-16**.The overall model is statistically significant, meaning that the predictors, as a group, are associated with sodium intake.

**Model 3 is statistically superior to Model 2, confirmed by the higher Adjusted R-squared. However, this marginal gain comes at a high cost as the model is significantly more complex (32 terms), and the poor significance of most individual interaction terms makes it extremely difficult to interpret for practical use.**

## Model-4 with variables up to 3 interaction terms
```{r}
model4 = lm(sodium ~ age + (gender + race + income_pov_cat)^3, data = food_model)
summary(model4)
xkabledply(model4, title = "Model for sodium with  variables upto 3 interactions")
xkablevif(model4)
```

**Model 4 was an attempt to add even more complexity by including additional interaction terms, aiming for the best predictive fit.This model includes all two-way and three-way interactions among gender, race, and income, in addition to age.**

**Coefficients Interpretation**

- **Main Effect Coefficients** The interpretation of main effects (Age, Gender) becomes more complex, as their effect is now dependent on the interaction terms.

- **Interaction Term Coefficients** This generated 49 new coefficients. These coefficients represent the additional increase or decrease in sodium intake for a specific race within a specific income category, beyond what the two main effects would predict.

**P-Values:**

- **Age  and genderFemale p-value < 2e-16**.This extremely small p-value indicates that age and genderFemale are highly significant predictors of sodium intake. We reject the null hypothesis that the age and genderFemale coefficient equals zero.

- **Race and Income Categories p-value** Almost all ethnic categories and income categories like are not statistically significant (p > 0.05), While only income_pov_cat300-400 is statistically significant (p = 0.00029).All of them lost statistical significance due to the inclusion of higher-order interactions, which absorb explanatory power.

- **Interaction Terms p-values** A small number of two-way interactions are statistically significant, but many are not.This suggests over-parameterization, where added complexity does not translate into substantial explanatory gains.

**Mutiple R-squared = 0.116 (1.16%)**. This model explains approximately 11.6% of the total variation in daily sodium intake. This model explains the highest percentage of variation among all models.

**Adjusted R-squared = 0.103**. The Adjusted R-squared score is likely lower than Model 3's, as the addition of non-significant terms results in an expense that exceeds the minor gain in fit.

**F-statistic p-value < 2.2e-16**.The overall model is statistically significant, due to the strong main effects (Age, Gender) carried over from Model 2.

**While Model 4 achieves the highest Multiple R-squared, it suffers from interpretability issues and limited marginal improvement. The large number of interaction terms increases variance and reduces clarity, making it less suitable for practical inference.**

# Model Testing

## Anova Test
```{r}
anova = anova(model1,model2,model3,model4)
anova
xkablesummary(anova)
```

The ANOVA results compare four nested linear regression models for sodium intake, each adding progressively more demographic and socioeconomic complexity.

**Model 2 vs. Model 1** - Adding gender, race, income, and time in the U.S. leads to a large and statistically significant reduction in residual sum of squares (F = 35.77, p < 2e-16). This indicates that demographic and socioeconomic variables substantially improve the model’s explanatory power beyond age and household size alone.

**Model 3 vs. Model 2** - Introducing race * income interaction terms results in a statistically significant but modest improvement in model fit (F = 1.62, p = 0.047). This suggests that the effect of income on sodium intake differs across racial groups, although the magnitude of improvement is limited.

**Model 4 vs. Model 3**- Adding higher-order (three-way) interactions does not significantly improve model fit at the alpha level (p = 0.097). This indicates diminishing returns from increased model complexity and potential overfitting.

**Model 2 provides a strong improvement over the baseline, while Model 3 offers additional insight into socioeconomic heterogeneity with minimal complexity increase. Model 4 does not justify its added complexity. Therefore, Model 3 represents the best balance between explanatory power, interpretability, and statistical justification for modeling determinants of sodium intake.**

## Best Model Visualizations

```{r}
#Extract and Prepare Model 2 Data
model2_data <- tidy(model2) %>%
  # Filter out the (Intercept) for cleaner visualization of predictors
  filter(term != "(Intercept)") %>% 
  # Create a significance label
  mutate(
    is_significant = p.value < 0.05,
    short_term = gsub("raceNon-Hispanic |income_pov_cat|gender", "", term) 
  )

#Create the P-value Bar Plot
pplot_model2 = ggplot(model2_data, aes(x = p.value, y = reorder(short_term, p.value), fill = is_significant)) +
  geom_bar(stat = "identity") +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red", size = 1) +
  geom_text(aes(label = round(p.value, 4)), 
            hjust = -0.1, 
            color = "black", 
            size = 7) +
  xlim(0, max(model2_data$p.value) * 1.1) +
  scale_fill_manual(values = c("TRUE" = "darkgreen", "FALSE" = "lightcoral"),
                    labels = c("Not Significant (p ≥ 0.05)", "Significant (p < 0.05)")) +
  labs(
    title = "P-values of Predictors in Model 2",
    x = "P-value",
    y = "Predictor Term",
    fill = "Significance"
  ) +
  theme(
    #Font sizes
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 8, face = "bold"),
    axis.text  = element_text(size = 8),
    legend.position = "bottom"
  )

#display the plot 
pplot_model2
```

This plot shows the p-values of the predictors of model2. This shows green bars for predictors like Age and Gender, indicating they are statistically significant, while red bars for predictors like race, income and time in us indicating they are not statistically significant.

```{r}
#Extract and Prepare Model 3 Data
pplott_model3 = model3_data <- tidy(model3) %>%
  # Filter out the (Intercept) for cleaner visualization of predictors
  filter(term != "(Intercept)") %>% 
  # Create a significance label
  mutate(
    is_significant = p.value < 0.05,
    short_term = gsub("raceNon-Hispanic |income_pov_cat|gender", "", term) 
  )

#Create the P-value Bar Plot
pplot_model3 = ggplot(model3_data, aes(x = p.value, y = reorder(short_term, p.value), fill = is_significant)) +
  geom_bar(stat = "identity") +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red", size = 1) +
  geom_text(aes(label = round(p.value, 4)), 
            hjust = -0.1, 
            color = "black", 
            size = 3) +
  xlim(0, max(model3_data$p.value) * 1.1) +
  scale_fill_manual(values = c("TRUE" = "darkgreen", "FALSE" = "lightcoral"),
                    labels = c("Not Significant (p ≥ 0.05)", "Significant (p < 0.05)")) +
  labs(
    title = "P-values of Predictors in Model 3",
    x = "P-value",
    y = "Predictor Term",
    fill = "Significance"
  ) +
  theme(
    #Font sizes
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 8, face = "bold"),
    axis.text  = element_text(size = 8),
    legend.position = "bottom"
  )

#display the plot 
pplot_model3
```

This plot shows the p-values of the predictors of model3.Most of the interaction terms added in Model 3 have high p-values, meaning they are not statistically significant individually.

**Since Model 3 is much harder to interpret without adding meaningful predictive power, we determined that Model 2 was the better choice for interpretability.**


```{r}
# Comprehensive residual plots
par(mfrow = c(2, 2)) 
plot(model3, main = "Diagnostic Plots for Model 3") 
```

The residual diagnostics indicate that several key linear regression assumptions are violated. 

**The Residuals vs. Fitted plot** shows a clear funnel-shaped pattern, with residual variance increasing as fitted sodium intake values rise. This indicates heteroscedasticity, meaning the model’s prediction errors are not constant across levels of the outcome.

**The Q–Q plot** reveals substantial deviations from the reference line, particularly in the upper tail. This suggests that the residuals are not normally distributed and are right-skewed, with more extreme high sodium values than expected under normality.

**The Residuals vs. Leverage plot** identifies a small number of observations with large residuals; however, none exhibit excessively high leverage or Cook’s distance. This indicates that while outliers exist, they are not highly influencing the estimated regression coefficients.

**Despite the clear assumption violations, Model 2 is choosen as the best untransformed option due to its superior interpretability and simplicity. These issues necessitate a data transformation (e.g Log-transformation) to produce a statistically robust model.**


# Transformations
```{r}
food_model$log_sodium = log(food_model$sodium)
```

To address the normality and heteroscedasticity issues we just identified, we applied a logarithmic transformation to our outcome variable.


## Histogram of Log-Transformed Sodium Intake
```{r}
hlog_sodium <- ggplot(food_model, aes(log(sodium))) +
  geom_histogram(
    bins = 40,
    fill = "chocolate",
    color = "chocolate4",
    alpha = 0.9
  ) +
  labs(
    title = "Distribution of Log-Transformed Sodium Intake",
    x = "log(Sodium)",
    y = "Frequency"
  ) +
  theme(
    plot.background = element_rect(fill = "antiquewhite", color = NA),
    panel.background = element_rect(fill = "antiquewhite", color = NA),
    panel.grid.major = element_line(color = "white"),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 20, face = "bold"),
    axis.text  = element_text(size = 20)
  )
# Display
hlog_sodium
```

This histogram shows the distribution of log-transformed sodium intake. After log transformation, the distribution is much more symmetric and bell-shaped, approximating a normal distribution much better.

## Linear Regression Model With log(sodium)
```{r}
model_log <- lm(log_sodium ~ age + gender + race + income_pov_cat + timeus + hhsize, data = food_model)
summary(model_log)
xkablevif(model_log)
```

**This model uses log(sodium intake) as the dependent variable to address skewness and heteroscedasticity observed in the untransformed models.The predictors include age, gender, race, income category, length of time in the U.S., and household size.**

**Coefficient Interpretation**

- **Intercept = 8.040911**, which represents the expected daily log sodium intake (in milligrams) when all variables are zero. While not directly meaningful substantively, it serves as the baseline from which all demographic effects are measured.

- **Age Coefficient = −0.0048**, Holding all other variables constant, each additional year of age is associated with approximately a 0.48% decrease in daily sodium intake. This suggests that older individuals tend to consume less sodium than younger individuals.This shows a negative effect on log sodium intake.

- **Gender(Female) Coefficient = -0.270793**, Being female is associated with approximately a 27% lower sodium intake compared to males, after controlling for all other factors. This is one of the largest effects in the model.So, Gender is a strong and highly significant predictor.

- **Race Non-Hispanic White Coefficients = 0.073** - Non-Hispanic White adults are predicted to consume approximately 7.62% more sodium than the reference group Hispanic.

- **Race Non-Hispanic Asian Coefficients = 0.147** - Non-Hispanic Asian adults are predicted to consume approximately 15.86% more sodium than the reference group Hispanic.

- **income_pov_cat500+ Coefficients = 0.183357** - Individuals in the highest income category are predicted to consume approximately 18.33% more sodium than the lowest income category (<100%, the reference).

- **income_pov_cat300-400 Coefficients = 0.190937** - Individuals in the 300-400% income category are predicted to consume approximately 19.09% more sodium than the lowest income category (<100%, the reference).

**P-Values:**

- **Highly Significant Predictors**: Age, genderFemale, and all income categories, are highly significant.

- **Significant Predictors (p < 0.05)**: raceNon-Hispanic White, raceNon-Hispanic Asian, raceOther Race, and income_pov_cat100-200 are statistically significant
predictors.

- **Non-Significant Predictors (p > 0.05)**: raceNon-Hispanic Black (p=0.585), hhsize (p=0.443), and the two timeus categories are not statistically significant predictors of sodium intake.

**Mutiple R-squared = 0.101 (10.1%)**. This model explains approximately 10.1% of the total variation in daily sodium intake.

**Adjusted R-squared = 0.0983** The adjusted value confirms that the demographic and socioeconomic factors explain slightly under 10% of the variance.

**F-statistic p-value < 2.2e-16**.The overall model is statistically significant, meaning that the predictors, as a group,predict log(sodium) intake.

**The model log(sodium) is the best model as it satisfies the underlying regression assumptions. It provides strong evidence that Gender, Age, Race, and Income-to-Poverty Ratio are the most significant demographic drivers of sodium intake, with females consuming less and older, higher-income, and certain non-Hispanic groups consuming more..**


## Comprehensive residual plots
```{r}
par(mfrow = c(2, 2)) 
plot(model_log, main = "Diagnostic Plots for Model log(sodium) ")
```

These plots for the final model with log(sodium intake) as the dependent variable indicate that the linear regression assumptions are substantially better satisfied compared to the untransformed models.

**Residuals vs Fitted**:The residuals are now randomly scattered around zero with a relatively constant spread across all fitted values.This indicates that the log transformation has effectively corrected the heteroscedasticity problem. Prediction errors are now more stable across low, medium, and high levels of sodium intake.

**Normal Q–Q Plot**:The residuals follow the theoretical normal line much more closely than in the untransformed specification.Although mild deviations remain in the extreme tails, the central portion of the distribution aligns well with the diagonal, suggesting that the normality assumption is reasonably satisfied.

**Residuals vs Leverage**: A small number of observations show relatively large standardized residuals; however, none approach the Cook’s distance threshold. This indicates that while some outliers remain, they do not exert disproportionate influence on the estimated regression coefficients.

**The log transformation successfully mitigates the primary assumption violations—heteroscedasticity and non-normality—present in the original models.**

In conclusion, while demographic variables explain only about a small amount of the variation in sodium consumption, we identified distinct, statistically significant patterns:

- **Women consume significantly less sodium than men.**

- **Older adults tend to consume less sodium than younger adults.**

- **Higher Income groups and Non-Hispanic White or Asian populations are associated with significantly higher sodium intake.**

These findings are crucial for public health, as they help identify exactly which groups are at higher risk for excessive sodium consumption.
