---
title: "Project2_WriteUp_Model 3 Logisitical Regression"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_knit$set(
  root.dir = "/Users/marlon/Documents/EDUCATION/GWU/GWU Courses/GWU Fall 25/Introduction to Data Science_Darcy Morris/ModelDataWhatWeEat"
)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)


```

### Model 3: Logistic Regression – Predicting a Healthy Diet
```{r}
# Load required libraries
library(arrow)
library(ggplot2)
library(ezids)
library(labelled)
library(gtsummary)
library(openxlsx)
library(dplyr)
library(tidyr)
library(vctrs)
library(tibble)

# Set working directory (adjust if needed)
getwd()

food = read_parquet("proc_data/demo_foods.parquet")



# Quick inspection of the data
head(food)
names(food)
```

# 1. Constructing the Healthy Diet Score (Gender-Specific)

We construct a healthy diet score ranging from 0–7, where each point corresponds to meeting one NIH dietary benchmark. Because recommended intake differs by gender, we apply gender-specific thresholds for calories, protein, fiber, and carbohydrates. A binary outcome variable (healthy_diet) is then created, where 1 indicates meeting at least 5 of the 7 criteria. Roughly 24% of respondents meet this threshold, indicating a class imbalance.

```{r}
# Create a gender-specific healthy diet score based on NIH benchmarks
food <- food %>%
  mutate(
    healthy_score = case_when(
      # Female thresholds (RIAGENDR == 2)
      RIAGENDR == 2 ~ (
        (DR1IKCAL_sum <= 2000) +
        (DR1ISODI_sum < 2300) +
        (DR1IPROT_sum >= 50) +
        (DR1ISUGR_sum <= 50) +
        (DR1ICARB_sum <= 275) +
        (DR1IFIBE_sum >= 28) +
        (DR1ITFAT_sum <= 78)
      ),
      # Male thresholds (RIAGENDR == 1)
      RIAGENDR == 1 ~ (
        (DR1IKCAL_sum <= 2500) +
        (DR1ISODI_sum < 2300) +
        (DR1IPROT_sum >= 56) +
        (DR1ISUGR_sum <= 50) +
        (DR1ICARB_sum <= 300) +
        (DR1IFIBE_sum >= 31) +
        (DR1ITFAT_sum <= 88)
      )
    ),
    # Healthy if meeting at least 5 out of 7 criteria
    healthy_diet = if_else(healthy_score >= 5, 1, 0)
  )

# Distribution of healthy vs unhealthy
table(food$healthy_diet)
prop.table(table(food$healthy_diet))
```



# 2. Recoding Demographic Variables

To simplify interpretation and avoid sparse categories, we regroup marital status and race variables into broader, more balanced groups. Hispanic respondents are combined into a single category, and marital status is collapsed into partnered, widowed/separated, and single.

```{r}
# Recode marital status
food <- food %>%
  mutate(
    marital_recode = case_when(
      DMDMARTZ %in% c(1, 6) ~ "Partnered",
      DMDMARTZ %in% c(2, 4) ~ "Wid/Separated",
      TRUE ~ "Single"
    )
  )

# Recode race/ethnicity
food <- food %>%
  mutate(
    race_recode = case_when(
      RIDRETH3 %in% c(1, 2) ~ "Hispanic",
      RIDRETH3 == 3 ~ "White",
      RIDRETH3 == 4 ~ "Black",
      TRUE ~ "Other"
    )
  )
```


# 3. Preparing the Modeling Dataset

Here we prepare the final modeling dataset by converting categorical variables to factors and removing observations with missing values. The final dataset includes demographic predictors commonly used in diet research.

```{r}
# Convert categorical variables to factors
food$healthy_diet   <- factor(food$healthy_diet)
food$RIAGENDR       <- factor(food$RIAGENDR)
food$race_recode    <- factor(food$race_recode)
food$marital_recode <- factor(food$marital_recode)
food$DMDEDUC2       <- factor(food$DMDEDUC2)

# Select variables for modeling
model_data <- food %>%
  select(
    healthy_diet,
    RIAGENDR,
    RIDAGEYR,
    race_recode,
    marital_recode,
    DMDEDUC2,
    INDFMPIR
  ) %>%
  drop_na()
```


# 4. Stepwise Logistic Regression and ANOVA Comparison

We estimate a series of nested logistic regression models and compare them using likelihood ratio tests. Education significantly improves model fit, income does not add explanatory power once gender-adjusted thresholds are used, race improves the model modestly, and marital status does not significantly improve fit.

```{r}
# Model A: age + gender
model_A <- glm(healthy_diet ~ RIAGENDR + RIDAGEYR,
               data = model_data, family = binomial)

# Model B: add education
model_B <- glm(healthy_diet ~ RIAGENDR + RIDAGEYR + DMDEDUC2,
               data = model_data, family = binomial)

anova(model_A, model_B, test = "Chisq")

# Model C: add income
model_C <- glm(healthy_diet ~ RIAGENDR + RIDAGEYR + DMDEDUC2 + INDFMPIR,
               data = model_data, family = binomial)

anova(model_B, model_C, test = "Chisq")

# Model D: add race
model_D <- glm(healthy_diet ~ RIAGENDR + RIDAGEYR + DMDEDUC2 +
                 INDFMPIR + race_recode,
               data = model_data, family = binomial)

anova(model_C, model_D, test = "Chisq")

# Model E: add marital status
model_E <- glm(healthy_diet ~ RIAGENDR + RIDAGEYR + DMDEDUC2 +
                 INDFMPIR + race_recode + marital_recode,
               data = model_data, family = binomial)

anova(model_D, model_E, test = "Chisq")
```

# 5. Final Model Selection

Following the ANOVA evaluation, we identify model D as the best model. 
Using a 0.5 cutoff, model D achieves relatively high accuracy (~76%) but extremely low sensitivity. This reflects the strong class imbalance (see above) and indicates that demographic variables alone are poor predictors of healthy diet status.


```{r}
model3 <- model_D
# Predicted probabilities
pred_prob3 <- predict(model3, type = "response")

# Classification using 0.5 cutoff
pred_class3 <- if_else(pred_prob3 >= 0.5, 1, 0)

# Confusion matrix
conf_mat3 <- table(pred_class3, model_data$healthy_diet)
conf_mat3

# Performance metrics
cutoff <- 0.5
TN <- conf_mat3["0","0"]
FP <- conf_mat3["1","0"]
FN <- conf_mat3["0","1"]
TP <- conf_mat3["1","1"]

metrics <- list(
  Cutoff_used = cutoff,
  Accuracy = (TP + TN) / (TP + TN + FP + FN),
  Sensitivity = TP / (TP + FN),
  Specificity = TN / (TN + FP),
  Precision = TP / (TP + FP)
)

metrics
```


# 7. Visualizations

The visualizations highlight the distribution of diet scores by gender and confirm that the model performs only slightly better than random guessing (AUC ≈ 0.60), reinforcing the conclusion that demographics alone cannot strongly predict diet quality.

```{r}
# Distribution of healthy diet scores by gender
ggplot(food, aes(x = healthy_score, fill = RIAGENDR)) +
  geom_histogram(position = "dodge", bins = 7, alpha = 0.85) +
  scale_fill_manual(values = c("#1f78b4", "#e31a1c"),
                    labels = c("Male", "Female")) +
  labs(
    title = "Distribution of Healthy Diet Scores by Gender",
    subtitle = "Scores represent number of NIH benchmarks met",
    x = "Healthy Diet Score",
    y = "Count",
    fill = "Gender"
  ) +
  theme_minimal(base_size = 14)
```

```{r}
library(pROC)
roc_obj <- roc(model_data$healthy_diet, pred_prob3)

ggplot(data.frame(
  tpr = roc_obj$sensitivities,
  fpr = 1 - roc_obj$specificities),
  aes(x = fpr, y = tpr)) +
  geom_line(color = "#1f78b4", linewidth = 1.2) +
  geom_abline(linetype = "dashed", color = "grey50") +
  labs(
    title = "ROC Curve for Healthy Diet Model",
    subtitle = paste("AUC =", round(auc(roc_obj), 3)),
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  theme_minimal(base_size = 14)
```


Overall, the goal was to predict whether someone meets a gender-adjusted healthy diet definition (at least 5 out of 7 NIH-style benchmarks) using demographics. Stepwise ANOVA suggests education improves model fit and race adds a small improvement, while income and marital status do not add much once the scoring is gender-adjusted. Model performance is limited (AUC ≈ 0.60), and with a 0.5 cutoff it predicts almost everyone as “unhealthy,” so accuracy is driven mostly by class imbalance rather than strong prediction. Although the model captures broad demographic trends, it performs poorly at individual-level prediction, reflecting both class imbalance and the limits of demographics alone. Overall, the results clearly show that most Americans do not meet NIH nutritional benchmarks.






