---
title: "Determinants of a Healthy Diet of U.S. adults 2021-2021"
author: "Team-1"
date: "2025-10-21"
output:
   rmdformats::readthedown:
     code_folding: hide
     number_sections: true
     toc_depth: 3
     toc_float: yes
#bibliography: references.bib
#csl: apa-5th-edition.csl     
---
```{r init, include=F}
library(arrow)
library(broom)
library(caret)
library(car)
library(corrplot)
library(data.table)
library(dplyr)
library(ezids)
library(ggplot2)
library(ggthemr)
library(ggthemes)
library(grid)
library(gridExtra)
library(gtsummary)
library(Hmisc)
library(knitr)
library(labelled)
library(lattice)
library(leaps)
library(modelr)
library(MuMIn)
library(openxlsx)
library(patchwork)
library(ROCR)
library(scales)
library(survey)
library(tibble)
library(tidyr)
library(tidyverse)
library(tidymodels)
library(treemapify)
library(vctrs)
library(wCorr)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  results = "markup")
options(scientific=T, digits = 3) 

# Set the root directory for all code chunks (important for reading files)
#knitr::opts_knit$set(root.dir = "/Users/akashreddyashanna/Desktop/Fall 2025/Intro to DS/DataScienceProject")
knitr::opts_knit$set(root.dir = "..")
```

```{r readDB}
#### Reading the parquet file
food <- read_parquet("proc_data/food_final_morevars.parquet")
```


# Introduction

Do our backgrounds influence how healthy we eat?
This project uses three models to reveal how demographic factors affect sugar consumption, sodium intake, and the probability of a healthy diet in U.S. adults (20+).

# SMART Questions

Here

# Data
**What We Eat in America Dietary Data? National Health and Nutrition Examination Survey (NHANES) August 2021-August 2023**

The “What We Eat in America” (WWEIA) dietary data, collected between August 2021 and August 2023, form part of the National Health and Nutrition Examination Survey (NHANES) conducted by the U.S. Centers for Disease Control and Prevention (CDC) in collaboration with the U.S. Department of Agriculture (USDA). This dataset provides nationally representative information on the types and quantities of foods and beverages consumed by individuals in the United States.[@u.s._department_of_agriculture_agricultural_research_service_what_2024]

Data were collected through telephone-administered 24-hour dietary recalls using the USDA’s 5-Step Automated Multiple-Pass Method (AMPM), a structured interview protocol designed to enhance accuracy and reduce underreporting. Each participant completed up to two dietary recalls, in which they reported all foods, beverages (including water), and supplements consumed from midnight to midnight on the previous day. [@williams_what_2025]

Nutrient values were derived from the USDA Food and Nutrient Database for Dietary Studies (FNDDS), which provides standardized nutrient composition data for thousands of food items. The 2021–2023 WWEIA cycle was based on an updated survey design and modified questionnaires introduced after the COVID-19 pandemic to reflect changes in food consumption patterns and supply chains.

The following dietary variables were included in the analysis:
Individual ID, USDA 8-digit food code, **Energy (kcal)**, **Protein (g)**, **Carbohydrates (g)**, **Total Sugars (g)**, **Dietary Fiber (g)**, **Total Fat (g)**, and **Sodium (mg)**.

To generate individual-level daily consumption data, all foods reported under the USDA 8-digit codes were aggregated by summing nutrient quantities per individual. This approach provides the total daily nutrient intake per participant, allowing for direct comparisons across individuals and demographic groups. Aggregation by total intake—rather than mean or per-meal intake—was chosen to capture complete daily nutritional exposure, consistent with standard NHANES analytical practices. Future analyses could disaggregate these data by food category (e.g., cereals or processed foods) to examine the consumption of items most affected by recent price increases.

Demographic information was obtained from NHANES demographic files and included:
**Individual ID**, **Gender**, **Age**, **Race/Hispanic origin (including Non-Hispanic Asian)**, **Education level (adults 20+)**, **Marital status**, **Household size**, and **Family income-to-poverty ratio**.

Following USDA recommendations, the analytic sample was restricted to adults aged 20 years and older to minimize bias from unequal age distributions across ethnic groups, as children’s food intake is inherently lower. Pregnant women were excluded because of their distinct nutritional requirements and dietary behaviors.

To ensure national representativeness, NHANES provides dietary day one sample weights (WTDRD1), accounting for survey design, oversampling, and non response. However, the results reported in this report do not included this weights in the process. Nonetheless, we identified that the distribution of demographic groups matches what Consensus Economics reports for the U.S. (See Exploratory Data Analysis for more details).

After merging the dietary and demographic components, the final analytic sample consisted of 4,760 individuals and 36 variables, representing U.S. adults between August 2021 and August 2023.

While the WWEIA-NHANES data provide a robust and nationally representative source for dietary assessment, several limitations should be acknowledged. First, the 24-hour dietary recall method relies on self-reported information, which is subject to recall bias and under reporting or over reporting, particularly for foods perceived as unhealthy. Second, each participant’s recall represents only one or two days of intake, which may not fully capture habitual dietary patterns or seasonal variations. Third, the use of aggregated nutrient totals does not account for meal timing or contextual factors, such as eating location or preparation method. Finally, subgroup analyses, such as by ethnicity or income, may still be affected by smaller sample sizes and sampling variability.

Despite these limitations, WWEIA-NHANES remains one of the most comprehensive and reliable datasets for examining dietary intake and its relationship to demographic and socioeconomic characteristics in the U.S. population.

# Model 1: Sugar intake

## Distribution of sugar

```{r sugardist}
#No missing in log sugar
df <- subset(food, is.finite(log_sugar) & !is.na(log_sugar))
survey_design <- svydesign(ids = ~1,weights = ~weight, data = food)

##### Distribution of Log of Sugar
## Weighted mean and sd for log_sugar
m_logsugar  <- with(df, weighted.mean(log_sugar, weight, na.rm = TRUE))
sd_logsugar <- with(df, {
  mu <- weighted.mean(log_sugar, weight, na.rm = TRUE)
  sqrt(weighted.mean((log_sugar - mu)^2, weight, na.rm = TRUE))
})

#Graph
ggplot(df, aes(x = log_sugar)) +
  geom_histogram(
    aes(
      y      = after_stat(density),
      weight = weight
    ),
    bins  = 30,
    alpha = 0.5,
    color = "#4DBBD5",
    fill  = "#4DBBD5"
  ) +
  geom_density(
    aes(weight = weight),
    color = "#00A087",
    size  = 1.2,
    alpha = 0.3,
    na.rm = TRUE
  ) +
  stat_function(
    fun  = dnorm,
    args = list(
      mean = m_logsugar,
      sd   = sd_logsugar
    ),
    color    = "#E64B35",
    size     = 1,
    linetype = "dashed"
  ) +
  labs(
    title    = "Distribution of Log Sugar Intake",
    subtitle = "Histogram with density curve and normal distribution overlay",
    x        = "Logarithm Sugar intake (g)",
    y        = "Density",
    caption  = "Note: Orange red dashed line represents the weighted normal distribution.\nCalculations using survey weights."
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.caption   = element_text(hjust = 0),
    legend.position = "bottom",
    strip.text      = element_text(face = "bold")
  )
```

### Sugar Outliers
```{r sugaroutliers}
### Outliers Log-Sugar
food_f = outlierKD2(df, log_sugar, rm = TRUE, boxplt = TRUE, qqplt = TRUE)

#160 observations
sum(is.na(food_f$log_sugar))

## Demographics for outliers
table(food_f$gender, is.na(food_f$log_sugar)) # Balanced
tapply(food_f$age, is.na(food_f$log_sugar), mean, na.rm = TRUE) #AVG higher for non-missing
tapply(food_f$incomepovline, is.na(food_f$log_sugar), mean, na.rm = TRUE) #AVG higher for non-missing
table(food_f$race, is.na(food_f$log_sugar)) #More Balanced

cat_plot_data <- food_f %>%
  mutate(sugar_missing = is.na(log_sugar)) %>%
  pivot_longer(cols = c(gender, race),
               names_to = "variable",
               values_to = "category") %>%
  group_by(variable, category) %>%
  summarise(prop_missing = mean(sugar_missing), .groups = "drop")

cat_plot_data %>%
  mutate(category = str_to_title(category)) %>%
  ggplot(aes(x = category, y = prop_missing)) +
  geom_col(fill = "#4DBBD5", alpha=0.7) + 
  facet_wrap(~ variable, scales = "free_x") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = NULL,
    y = "Proportion Missing",
    title = "Proportion of Log Sugar Missing by Category"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1),
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )

num_plot_data <- food_f %>%
  mutate(sugar_missing = ifelse(is.na(log_sugar), "Sugar missing", "Sugar not missing")) %>%
  pivot_longer(cols = c(age, incomepovline),
               names_to = "variable",
               values_to = "value") %>%
  group_by(variable, sugar_missing) %>%
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop")

ggplot(num_plot_data,
          aes(x = sugar_missing, y = mean_value, fill = sugar_missing)) +
  geom_col(alpha=0.7) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    x = "Sugar missing status",
    y = "Mean value",
    title = "Mean of Numeric Variables by Sugar Missing Status"
  ) +
  guides(fill = "none") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(hjust = 1),
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )
```

## Independent Variables

```{r}
##### Scatterplot of Sugar and Number of Snacks
ct <- cor.test(df$numsnacks, df$log_sugar, use = "complete.obs")
r_val <- unname(ct$estimate)   # Pearson r
p_val <- ct$p.value            # p-value
lab_text <- paste0("r = ", round(r_val, 3), "\n p = ", signif(p_val, 3))

ggplot(df, aes(x = numsnacks, y = log_sugar)) +
  geom_point(color = "#4DBBD5", alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1, color = "#E64B35") +
  annotate(
    "text", x = Inf, y = 0, hjust  = 1.1, vjust  = 1.5, label  = lab_text, size   = 5
  ) +
  labs(
    title = "Sugar Intake vs. Number of snacks",
    x = "Number of Snacks",
    y = "Log Sugar Intake (gm)"
  ) +
  theme_minimal(base_size = 18) +
  theme(
    plot.title = element_text( hjust = 0.5, face = "bold", size = 14),
    axis.title.x = element_text(margin = margin(t = 8)),
    axis.title.y = element_text(margin = margin(r = 8))
  )

```


```{r}
##### Scatterplot of Sugar and Number of Meals at Home
ct <- cor.test(df$numhomemeals, df$log_sugar, use = "complete.obs")
r_val <- unname(ct$estimate)   # Pearson r
p_val <- ct$p.value            # p-value
lab_text <- paste0("r = ", round(r_val, 3), "\n p = ", signif(p_val, 3))

ggplot(df, aes(x = numhomemeals, y = log_sugar)) +
  geom_point(color = "#4DBBD5", alpha = 0.5, size  = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1, color = "#E64B35") +
  annotate(
    "text", x = Inf, y = 0, hjust  = 1.1, vjust  = 1.5, label  = lab_text, size   = 5
  ) +
  labs(
    title = "Sugar Intake vs. Number of Meals at Home",
    x     = "Number of of Meals at Home", 
    y     = "Log Sugar Intake (gm)") +
  theme_minimal(base_size = 18) +
  theme(
    plot.title   = element_text(hjust = 0.5,face  = "bold",size  = 14),
    axis.title.x = element_text(margin = margin(t = 8)),
    axis.title.y = element_text(margin = margin(r = 8))
  )
```

```{r}
##### Scatterplot of Sugar and Age
ct <- cor.test(df$age, df$log_sugar, use = "complete.obs")
r_val <- unname(ct$estimate)   # Pearson r
p_val <- ct$p.value            # p-value
lab_text <- paste0("r = ", round(r_val, 3), "\n p = ", signif(p_val, 3))

ggplot(df, aes(x = age, y = log_sugar)) +
  geom_point(color = "#4DBBD5", alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1, color = "#E64B35") +
  annotate(
    "text", x = Inf, y = 0, hjust  = 1.1, vjust  = 1.5, label  = lab_text, size   = 5
  ) +  
  labs(
    title = "Sugar Intake vs. Age",
    x = "Age",
    y = "Log Sugar Intake (gm)"
  ) +
  theme_minimal(base_size = 18) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title.x = element_text(margin = margin(t = 8)),
    axis.title.y = element_text(margin = margin(r = 8))
  )
```

```{r}
my_colors <- c("<100"    = "#7AEBD1",
               "100-200"      = "#27B291",
               "200-300"      = "#12705A",
               "300-400"      = "#124E70",
               "400-500"      = "#277EB0",
               "500+" = "#6FC2F2",
               "NA" = "gray")


# ---- Prepare data: weighted sugar + within-variable shares ----
df_plot <- df %>%
  group_by(income_pov_cat) %>%
  summarise(
    weighted_n     = sum(weight, na.rm = TRUE),
    weighted_sugar = weighted.mean(log_sugar, weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    prop  = weighted_n / sum(weighted_n),      # share within each variable
    share_label = percent(prop, accuracy = 0.1)
  ) %>%
  ungroup()
scale_factor <- max(df_plot$weighted_sugar, na.rm = TRUE) / 
  max(df_plot$prop,           na.rm = TRUE)

df_plot <- df_plot %>%
  mutate(prop_scaled = prop * scale_factor)

# ---- Plot: ----
ggplot(df_plot, aes(x = income_pov_cat)) +
  geom_col(aes(y = weighted_sugar, fill = income_pov_cat, alpha = 0.4)) +
  geom_point(aes(y = prop_scaled), size = 3, color = "black") +
  geom_text(aes(y = prop_scaled, label = share_label),
            vjust = -0.7, size = 3.5) +
  scale_fill_manual(values = my_colors) +
  
  scale_y_continuous(
    name = "Average sugar consumption (grams)",
    breaks = seq(0, max(df_plot$weighted_sugar) * 1.1, by = 10),
    
    expand = expansion(mult = c(0, 0.1)),
    
    sec.axis = sec_axis(
      ~ . / scale_factor,
      name = "Percentage of people in each category",
      breaks = seq(0, 1, by = 0.05),   # every 5%
      labels = percent_format()
    )
  ) +
  
  labs(
    title = "Average Sugar Consumption and Sample Shares by Income",
    caption = "Source:Author's calculations using WWEIA 2021-2023. \nNote: Calculations using survey weights.",
    x = "",
    fill = ""
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.caption = element_text(hjust = 0, size = 10, margin = margin(t = 10))
  )

```

```{r}
my_colors <- c("gender"    = "#4DBBD5",
               "race"      = "#E64B35",
               "education" = "#00A087")


# ---- Prepare data: weighted sugar + within-variable shares ----
df_plot <- food %>%
  pivot_longer(
    cols = c(gender, race, education),
    names_to = "variable",
    values_to = "category"
  ) %>%
  group_by(variable, category) %>%
  summarise(
    weighted_n     = sum(weight, na.rm = TRUE),
    weighted_sugar = weighted.mean(sugar, weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(variable) %>%
  mutate(
    prop  = weighted_n / sum(weighted_n),      # share within each variable
    share_label = percent(prop, accuracy = 0.1)
  ) %>%
  ungroup() %>%
  mutate(full_cat = paste(variable, category, sep = ": "))

# ---- Rescale share so dots can be plotted on sugar scale ----
scale_factor <- max(df_plot$weighted_sugar, na.rm = TRUE) / 
  max(df_plot$prop,           na.rm = TRUE)

df_plot <- df_plot %>%
  mutate(prop_scaled = prop * scale_factor)

# ---- Plot----
ggplot(df_plot, aes(x = category)) +
  geom_col(aes(y = weighted_sugar, fill = variable, alpha = 0.4)) +
  geom_point(aes(y = prop_scaled), size = 3, color = "black") +
  geom_text(aes(y = prop_scaled, label = share_label),
            vjust = -0.7, size = 3.5) +
  scale_fill_manual(values = my_colors) +
  scale_y_continuous(
    name = "Average sugar consumption",
    expand = expansion(mult = c(0, 0.1)),
    sec.axis = sec_axis(
      ~ . / scale_factor,
      name   = "Share of people within each category",
      labels = percent_format()
    )
  ) +
  labs(
    title = "Average Sugar Consumption and Category Shares (Weighted)",
    x = "",
    fill = ""
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r}
my_colors <- c("timeus"    = "#4DBBD5",
               "militar"      = "#E64B35",
               "maritalstatus" = "#00A087")


# ---- Prepare data: weighted sugar + within-variable shares ----
df_plot <- food %>%
  pivot_longer(
    cols = c(timeus, militar, maritalstatus),
    names_to = "variable",
    values_to = "category"
  ) %>%
  group_by(variable, category) %>%
  summarise(
    weighted_n     = sum(weight, na.rm = TRUE),
    weighted_sugar = weighted.mean(sugar, weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(variable) %>%
  mutate(
    prop  = weighted_n / sum(weighted_n),      # share within each variable
    share_label = percent(prop, accuracy = 0.1)
  ) %>%
  ungroup() %>%
  mutate(full_cat = paste(variable, category, sep = ": "))

# ---- Rescale share so dots can be plotted on sugar scale ----
scale_factor <- max(df_plot$weighted_sugar, na.rm = TRUE) / 
  max(df_plot$prop,           na.rm = TRUE)

df_plot <- df_plot %>%
  mutate(prop_scaled = prop * scale_factor)

# ---- Plot----
ggplot(df_plot, aes(x = category)) +
  geom_col(aes(y = weighted_sugar, fill = variable, alpha = 0.4)) +
  geom_point(aes(y = prop_scaled), size = 3, color = "black") +
  geom_text(aes(y = prop_scaled, label = share_label),
            vjust = -0.7, size = 3.5) +
  scale_fill_manual(values = my_colors) +
  scale_y_continuous(
    name = "Average sugar consumption",
    expand = expansion(mult = c(0, 0.1)),
    sec.axis = sec_axis(
      ~ . / scale_factor,
      name   = "Share of people within each category",
      labels = percent_format()
    )
  ) +
  labs(
    title = "Average Sugar Consumption and Category Shares (Weighted)",
    x = "",
    fill = ""
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r}
summary(aov(log_sugar ~ incomepovline, data=df))
summary(aov(log_sugar ~ gender, data=df))
summary(aov(log_sugar ~ race, data=df))
summary(aov(log_sugar ~ education, data=df))
summary(aov(log_sugar ~ timeus, data=df))
summary(aov(log_sugar ~ militar, data=df))
summary(aov(log_sugar ~ maritalstatus, data=df))
```

### Correlation
```{r}
########### Correlation Variables
dffilt=df[c("log_sugar","gender", "education", "race", "timeus",
            "age", "income_pov_cat", "maritalstatus", "militar",
            "numsnacks", "numhomemeals", "dayweek")]

pairs(dffilt)

cornum=cor(dffilt[c("log_sugar", "age", "numsnacks", "numhomemeals")])
corrplot(cornum)
```

## Linear Model

Log sugar on xxx

### Best Model using LEAPS
```{r}
##### USE SUBSECT LEAPS
subfit <- regsubsets(
  log_sugar ~ ., data = dffilt,
  nvmax = 10,           # max number of predictors allowed
  method = "exhaustive" # or "forward", "backward", "seqrep"
)

subsum <- summary(subfit)

# Look at criteria by subset size
subsum$adjr2   # adjusted R^2
subsum$cp      # Mallows' Cp
subsum$bic     # BIC

best_r2 <- which.max(subsum$adjr2)
best_bic <- which.max(subsum$bic)

best_coefs <- coef(subfit, best_r2)
best_vars  <- names(best_coefs)[-1]  # drop "(Intercept)"
best_vars
best_coefs
```

### Best Model ANOVA comparison
```{r}
##### MANUAL IDENTIFICATION
#Benchmark
fit1 <- glm(log_sugar ~ numsnacks + numhomemeals, data = df)
# + gender
fit2 <- glm(log_sugar ~ numsnacks + numhomemeals + gender, data = df)
anova(fit1,fit2)
# + education
fit3 <- glm(log_sugar ~ numsnacks + numhomemeals + gender + education, data = df)
anova(fit1,fit2, fit3)
# + age
fit4 <- glm(log_sugar ~ numsnacks + numhomemeals + gender + education + age, data = df)
anova(fit1,fit2, fit3, fit4)
# + age & age^2
fit5 <- glm(log_sugar ~ numsnacks + numhomemeals + gender + education + age + I(age^2), data = df)
anova(fit1,fit2, fit3, fit4, fit5)

anova_res <- anova(fit1,fit2, fit3, fit4, fit5, test = "F")
anova_df <- as.data.frame(anova_res)
added_terms <- c(NA, "gender", "education", "age", "age^2")
result_table <- data.frame(Model = paste0("Model ", 1:5),
  Added_Term = added_terms,
  P_Value = c(NA, anova_df$`Pr(>F)`[-1]))

result_table


### New Benchmark
fit1 <- glm(log_sugar ~ numsnacks + numhomemeals + gender + education, data = df)

#+race
fit2 <- glm(log_sugar ~ numsnacks + numhomemeals + gender + education + race, data = df)
anova(fit1,fit2)

#+timeus
fit3 <- glm(log_sugar ~ numsnacks + numhomemeals + gender + education + race +timeus, data = df)
anova(fit1,fit2, fit3)

#maritalstatus
fit4 <- glm(log_sugar ~ numsnacks + numhomemeals + gender + education + race +timeus+maritalstatus, data = df)
anova(fit1,fit2, fit3, fit4)

anova_res <- anova(fit1,fit2, fit3, fit4, test = "F")
anova_df <- as.data.frame(anova_res)
added_terms <- c(NA, "race", "timeus", "maritalstatus")
result_table <- data.frame(Model = paste0("Model ", 1:4),
                           Added_Term = added_terms,
                           P_Value = c(NA, anova_df$`Pr(>F)`[-1]))

result_table

### New Benchmark
fit1 <- glm(log_sugar ~ numsnacks + numhomemeals + gender + education + race +timeus, data = df)
#+militar
fit2 <- glm(log_sugar ~ numsnacks + numhomemeals + gender + education + race +timeus + militar, data = df)
anova(fit1,fit2)
#+dayweek
fit2 <- glm(log_sugar ~ numsnacks + numhomemeals + gender + education + race +timeus + dayweek, data = df)
anova(fit1,fit2)
#+income
fit2 <- glm(log_sugar ~ numsnacks + numhomemeals + gender + education + race +timeus + income_pov_cat, data = df)
tempf1=glm(log_sugar ~ numsnacks + numhomemeals + gender + education + race +timeus, data = df[!is.na(df$income_pov_cat), ])
anova(tempf1,fit2)
```

### Best Model
```{r}
dffinal=df[!is.na(df$income_pov_cat), ]

fitfinal=glm(log_sugar ~ numsnacks + numhomemeals + gender + education + race +timeus + income_pov_cat, data = dffinal)
summary(fitfinal)
```

CoefPlot
```{r}
######### SELECTED MODEL
## Coefplot
coef_df <- tidy(fitfinal, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # significance stars
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    ),
    coef_label = sprintf("%.3f%s", estimate, sig),
    term = gsub("_", " ", term),
    term = factor(term, levels = rev(unique(term)))
  )

ggplot(coef_df, aes(x = term, y = estimate)) +
  geom_point(size = 3, color = "#4DBBD5", alpha=0.5) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                width = 0.2, color = "#4DBBD5", alpha=0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5, color="#E64B35") +
  geom_text(aes(label = coef_label), vjust=-0.5,
            hjust = 0, size = 3.5) +
  coord_flip() +
  labs(
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

Model Fit
```{r}
r.squaredGLMM(fitfinal)
#Null
pchisq(2670.1, 4237, lower.tail=F)
#Fitt
pchisq(2398.4, 4219, lower.tail=F)
#Best?
pchisq(2670.1-2398.4, 4237-4219, lower.tail=F)
# Saving the predictions:
dfwithpred <- add_predictions(dffinal,fitfinal)
#Plot predictions vs real value
ggplot(dfwithpred, aes(x = log_sugar, y = pred)) +
  geom_point(alpha = 0.5, color = "#4DBBD5", size = 2) +
  geom_line(aes(pred), colour = "#E64B35", size = 1.2) +
  labs(
    x = "Observed log(Sugar Intake)",
    y = "Predicted log(Sugar Intake)",
    title = "Observed vs. Predicted Sugar Intake"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold")
  )

## Exploring Model
par(mfrow = c(1,1))
plot(fitfinal, ask=FALSE)
```

Multicollinearity
```{r}
vif(fitfinal)
```

IT HAS HETEROSCEDASTICITY AND THE ERRORS DIFFER FROM NORMAL AT THE TAIL