---
title: "NHANES EDA — Demographics & Nutrients"
author: "Team"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
    df_print: paged
---

```{r}
# knitr / plot options for clean writeup
knitr::opts_chunk$set(
  echo = TRUE,       # show code
  message = FALSE,   # hide package startup messages
  warning = FALSE,   # hide warnings in the doc
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

```{r setup, include=FALSE}
# Set knit options (affects all chunks)
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

# Set the root directory for all code chunks (important for reading files)
knitr::opts_knit$set(root.dir = "/Users/marlon/DataScienceProject/proc_data")

# Load required libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(arrow)       # to read parquet files
library(gtsummary)   # for summary tables
library(labelled)    # for variable labels if present
library(patchwork)
library(ezids)
library(labelled)
library(openxlsx)
library(vctrs)
library(tibble)
library(broom)
```

```{r}
# NOTE: the file should be in the working dir of the proc data or adjust the path
getwd()
# this is the processed file we created earlier in the pipeline
food <- read_parquet("demo_foods.parquet")

# quick peek to make sure it loaded 
dplyr::glimpse(food)
```
**Part I: EDA**
#Demographics (quick summaries)
```{r}
# selecting a few categorical vars we care about
# gives us a nice overview of N and key characteristics
demo_cat_tbl <- food |>
  select(RIAGENDR, RIDRETH3, DMDEDUC2, DMDMARTZ) |>
  tbl_summary(
    type = list(
      everything() ~ "categorical"
    ),
    missing = "ifany"   # show missing if present
  ) |>
  bold_labels()

demo_cat_tbl
```
**2) Continious Variables**
```{r}
# summarize a few continuous vars (mean/sd/min/max + missing)
library(gtsummary)

demo_cont_tbl <- food |>
  select(RIDAGEYR, DMDHHSIZ, INDFMPIR) |>
  tbl_summary(
    type = list(everything() ~ "continuous2"),
    statistic = all_continuous() ~ c(
      "Observations = {N_obs}",
      "Missing = {N_miss}",
      "Mean (SD) = {mean} ({sd})",
      "Min, Max = {min}, {max}"
    ),
    digits = everything() ~ 2
  ) |>
  bold_labels()
```

**3) Nutrients**
```{r}
# relabel a few key nutrient totals (Day 1 sums) for nicer plot titles
nutrient_df <- food |>
  rename(
    `Calories (kcal)`   = DR1IKCAL_sum,
    `Protein (g)`       = DR1IPROT_sum,
    `Carbohydrates (g)` = DR1ICARB_sum,
    `Sugars (g)`        = DR1ISUGR_sum,
    `Fiber (g)`         = DR1IFIBE_sum,
    `Total Fat (g)`     = DR1ITFAT_sum,
    `Sodium (mg)`       = DR1ISODI_sum
  )
# simple sanity check of the columns exist
names(nutrient_df)
```
**4) Distribution & assumptions**
Histogram for Protein overall
```{r}
p_hist <- ggplot(nutrient_df, aes(`Protein (g)`)) +
  geom_histogram(
    bins = 40,
    fill = "chocolate",
    color = "chocolate4",
    alpha = 0.9
  ) +
  labs(
    title = "Distribution of Day 1 Protein Intake",
    x = "Protein (g)",
    y = "Count"
  ) +
  theme_minimal(base_size = 13)

p_hist

# (optional) also save to file for slides
ggsave("output/EDA_hist_prot.png", p_hist, width = 8, height = 6, dpi = 300)

```
Boxplot for Protein overall
```{r}
p_box <- ggplot(nutrient_df, aes(y = `Protein (g)`)) +
  geom_boxplot(
    fill = "chocolate",
    color = "chocolate4",
    outlier.alpha = 0.5
  ) +
  labs(
    title = "Boxplot: Day 1 Protein Intake",
    x = "",
    y = "Protein (g)"
  ) +
  theme_minimal(base_size = 13)

p_box
ggsave("output/EDA_box_prot.png", p_box, width = 8, height = 6, dpi = 300)
```
Q–Q plot Protein overall
```{r}
p_qq <- ggplot(nutrient_df, aes(sample = `Protein (g)`)) +
  stat_qq(alpha = 0.5) +
  stat_qq_line(linewidth = 0.7) +
  labs(
    title = "Q–Q Plot: Day 1 Protein Intake",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_minimal(base_size = 13)

p_qq
ggsave("output/EDA_qq_prot.png", p_qq, width = 8, height = 6, dpi = 300)

```
**5) Protein by Gender**
```{r}
# make a simple gender label (1 = Male, 2 = Female in NHANES)
gender_df <- nutrient_df |>
  mutate(gender = case_when(
    RIAGENDR == 1 ~ "Male",
    RIAGENDR == 2 ~ "Female",
    .default = NA_character_
  ))

prot_gender_summary <- gender_df |>
  filter(!is.na(gender)) |>
  group_by(gender) |>
  summarise(
    n = n(),
    mean = mean(`Protein (g)`, na.rm = TRUE),
    sd = sd(`Protein (g)`, na.rm = TRUE),
    se = sd / sqrt(n),
    ci_low = mean - 1.96 * se,
    ci_high = mean + 1.96 * se,
    .groups = "drop"
  )

prot_gender_summary
```

```{r protein-ci-plot, echo=TRUE, fig.width=8, fig.height=6}
# Plot mean protein intake by gender with 95% confidence intervals
p_ci <- ggplot(prot_gender_summary, aes(x = gender, y = mean, color = gender)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high),
                width = 0.12, linewidth = 0.8) +
  labs(
    title = "Mean Protein Intake (95% CI) by Gender",
    x = "",
    y = "Protein (g)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

# Display the plot in the knitted document
p_ci
```
###FIX Tomorrow#####
```{r hist-protein-gender, fig.width=7, fig.height=5}
# Overlapping histograms of protein intake by gender
hprot_gender <- ggplot(food, aes(x = DR1IPROT_sum, fill = RIAGENDR)) +
  geom_histogram(
    breaks = seq(0, 500, by = 10),
    color = "black",
    alpha = 0.6,
    position = "identity"
  ) +
  # Set blue for males and red for females
  scale_fill_manual(values = c("Male" = "steelblue3", "Female" = "indianred2")) +
  labs(
    title = "Distribution of Daily Protein Intake by Gender",
    x = "Protein Intake (grams)",
    y = "Frequency",
    fill = "Gender"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    legend.position = "top"
  )

# Display the plot
hprot_gender
```

**EDA Summary Protein by Gender***
Summarizing the findings in four graphs using patchwork

```{r protein-multipanel, echo=TRUE, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
# packages (if not already loaded)


# --- 1) tidy data -----------------------------------------------------------
# RIAGENDR: 1 = Male, 2 = Female; DR1IPROT_sum = total protein (g)
prot_df <- food %>%
  transmute(
    gender = factor(if_else(RIAGENDR == 1, "Male", "Female"),
                    levels = c("Male", "Female")),
    protein = DR1IPROT_sum
  ) %>%
  filter(!is.na(gender), !is.na(protein))

# --- 2) compute mean + 95% CI (for point+CI panel) -------------------------
prot_gender_summary <- prot_df %>%
  group_by(gender) %>%
  summarise(
    n   = n(),
    mean = mean(protein),
    sd   = sd(protein),
    se   = sd / sqrt(n),
    tcrit = qt(0.975, df = n - 1),
    ci_low  = mean - tcrit * se,
    ci_high = mean + tcrit * se,
    .groups = "drop"
  )

# --- 3) four complementary views -------------------------------------------

# A) Mean with 95% CI (clean effect view)
p_ci <- ggplot(prot_gender_summary, aes(x = gender, y = mean, color = gender)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.15, linewidth = 0.9) +
  labs(title = "Mean Protein (95% CI) by Gender", x = NULL, y = "Protein (g)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

# B) Boxplot + mean dots (distribution + center)
p_box <- ggplot(prot_df, aes(x = gender, y = protein, fill = gender)) +
  geom_boxplot(outlier.alpha = 0.25) +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 2.5, fill = "white") +
  labs(title = "Distribution (Boxplot) + Mean", x = NULL, y = "Protein (g)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

# C) Density overlap (shape of distributions)
p_dens <- ggplot(prot_df, aes(x = protein, fill = gender)) +
  geom_density(alpha = 0.35) +
  labs(title = "Density Overlap", x = "Protein (g)", y = "Density") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")

# D) Violin + jitter (spread + individual points sample)
set.seed(42)
p_viol <- ggplot(prot_df %>% sample_n(min(1500, n())),  # subsample for readability
                 aes(x = gender, y = protein, fill = gender)) +
  geom_violin(trim = TRUE, alpha = 0.7) +
  geom_jitter(width = 0.15, alpha = 0.25, size = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 2.5, fill = "white") +
  labs(title = "Violin + Jitter (Subset)", x = NULL, y = "Protein (g)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

# --- 4) assemble into one figure -------------------------------------------
(p_ci | p_box) / (p_dens | p_viol)
```


**Part II: hypothesis testing** 
Do Women Eat Less Protein Than Men?

In this section, we statistically test whether the average daily protein intake differs by gender, specifically if **women consume less protein** than men.  
We’ll visualize the data first, then run a t-test and confirm results with an ANOVA.

```{r setup-hypothesis, message=FALSE, warning=FALSE}

# Convert gender to readable factor
food$RIAGENDR <- factor(food$RIAGENDR,
                        levels = c(1, 2),
                        labels = c("Male", "Female"))

```

**T-Test: Testing Mean Difference Between Men and Women**
```{r}
tres <- t.test(DR1IPROT_sum ~ RIAGENDR,
               data = food,
               alternative = "greater") # one-sided test
tres
```
**Tidy Summary Table**
```{r}
tres_tidy <- tidy(tres) %>%
  mutate(across(where(is.numeric), round, 2)) %>%
  rename(
    "Mean Male" = estimate1,
    "Mean Female" = estimate2,
    "Mean Difference (Male - Female)" = estimate,
    "t Statistic" = statistic,
    "p Value (one-sided)" = p.value,
    "CI Lower" = conf.low,
    "CI Upper" = conf.high
  ) %>%
  select(`Mean Male`, `Mean Female`, `Mean Difference (Male - Female)`,
         `CI Lower`, `CI Upper`, `t Statistic`, `p Value (one-sided)`)

tres_tidy
```
**ANOVA Test**

```{r}
anovares <- aov(DR1IPROT_sum ~ RIAGENDR, data = food)
summary(anovares)
```

